Galaxy image statistical generator: how to be confident?

### 2024-09-10 paper/main.tex
more on flows
diffusion model

The structure of the transformation (e.g. Eq.~\ref{eq-NVP_forward}) is a single split of the $\bm{z}$ vector coordinates in two parts: the first part $(z_{\leq s})$ is not transformed while the transformation of the second part ${z_{>s}})$ uses a function depending only on the first part. Notice that the conditioner functions do depends only on $(z_{\leq s})$ for both the forward or the backward transformations. This single split schema is illustrated on Figure \ref{fig-nvp}. 
%
\begin{figure}
\centering
\includegraphics[width=0.45\columnwidth]{fig-nvp.pdf}
\includegraphics[width=0.45\columnwidth]{fig-nvp-inverse.pdf}
\caption{Affine coupling layer schema of forward and backward directions (Eqs.~\ref{eq-NVP_forward},\ref{eq-NVP_backward}) with a single split mechanism (figure inspired by \cite{Papamakarios2021}).}
\label{fig-nvp}
\end{figure}
%
This splitting schema implemented in a so-called \textit{coupling layer} makes the flow-based model computationally efficient and using deep neural networks, one can achieve complex $(\bm{\alpha},\bm{\beta})$ functions. Now, designers of complete architectures can compose different \textit{coupling layers} by reversing $\bm{z}$ elements between layers to ensure that all elements are transformed at the end of the set of operations. An generalisation of any permutation ordering of the inputs can be performed using a $1\times 1$ convolution layer. \cite{2016arXiv160508803D} improves the modelling using a multi-scale approach which would be prohibitive to describe here, and \cite{Kingma2018} use all these extensions to build the \texttt{Glow} model whose schema is displayed on Figure \ref{fig-Glow-archi}.
\begin{figure}
    \centering
    \vspace{-0cm}
    \begin{subfigure}[b]{0.45\textwidth}
    \begin{center}
        \centering
        \includegraphics[width=.7\textwidth]{fig-Glow-1.pdf}
        \caption{One step of Glow flow.}
    \end{center}
    \end{subfigure}%
    \vspace{5mm}
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \includegraphics[width=.99\textwidth]{fig-Glow-2.pdf}
        \caption{Multi-scale architecture \citep{2016arXiv160508803D}.}
    \end{subfigure}
    \caption{Figure from \citep{Kingma2018}. The flow step is composed of a scale and bias layer with data dependent initialization (\texttt{actnorm}), a $1\times 1$ invertible convolution layer and an affine coupling layer with $(\bm{\alpha},\bm{\beta})$ issued from a shallow convolutional neural network. Both $\bm{x}$ and $\bm{y}$ are tensors of
shape $[h \times w \times c]$ with spatial dimensions $(h, w)$ and channel dimension $c$ ans $(i, j)$ denote
spatial indices into tensors $\bm{x}$ and $\bm{y}$.  This flow step is embedded in a multi-scale  architecture which has $K$ flow steps and a number of levels $L$ \citep{2016arXiv160508803D}. The squeezing operation reduces a $s \times s \times c$ tensor  into a $s/2\times s/2 \times 4c$ tensor which mix spatial and channel components. At each scale, each channel image is divided into $2\times 2$ patches ($s=2$) and each patch is squeezed. }
    \label{fig-Glow-archi}
\end{figure}
 %
\subsection{Score based diffusion models}
%
To generate samples from $p(\bm{x})$, we can proceed as follows \citep[e.g][]{Chang2023,LinYang2023}. Starting from a sample $\bm{x}_0$ assumed to be drawn from the distribution $p_0(\bm{x})=p(\bm{x})$, we progressively transform it until the  distribution is as simple as for instance $\mathcal{N}(\bm{z},\bm{0},\bm{1})$.  The idea is then to reverse the process (the \textit{backward} mode) by generating a new sample from a new instance of noise. This transfer mapping looks very similar in concept as normalizing flows. Their latent spaces have the same dimension as the data space, but the main difference consists in the fact that the later are deterministic by essence while the former are stochastic.

@@

Notice that in a VAE architecture beyond the information compression,  the transformation from $\bm{x}_0$ to $\bm{z}$ is performed by an \textit{encoder} whose implementation is typically a deep neural networks with trained parameters. This is not the case in the above stochastic process where only noise is added to the original signal (e.g. galaxy image). In the same spirit, the VAE \textit{decoder} or the inverted flows are replaces by stochastic differential equation (damped Langevin)
\begin{equation}
dx_{T-t} = (x_{T-t} + 2\nabla_x \log p_{T-t}(x_{T-t}))dt + \sqrt{2}\ dB_t \quad t\in[0,T]
\label{eq-backward-diffusion}
\end{equation}
where $\nabla_x \log p_{t}(\bm{x}_{t})$ is the \textit{score} of the probability density at time $t$. This backward process is the $\bm{x}=\bm{x}_0$ generative direction from $\mathcal{N}(\bm{z},\bm{0},\bm{1})$ sampling.

The key problem is how to obtain the score. One can think about two possible approaches: either we have a model for $p_t$ \citep[e.g.][]{Guth2022b,Lempereur2024} that can exploit regularity patterns and assumptions about correlations to provide a constrained parametric framework for learning the score by score matching \citep{hyvarinen2005a}, or we use a \textit{denoising} neural network as introduced by \cite{Bengio2013} and then \cite{Sohl-Dickstein2015,Ho2020} and recently studied from a mathematical point of view by \citep{kadkhodaie2024generalization}, as well as used in cosmology such as the generation of galaxy images \citep{smith2021}, the generation of 21cm luminosity temperature maps \citep{Zhao2023} and super-resolved large-scale cosmic structure predictions \citep{Schanz2023} and sampling of the high-dimensional Bayesian posterior of the weak lensing mass-mapping problem \citep{Remy2023}.

Denoising diffusion stochastic models have certain advantages compared to GAN....

They can jointly be interplayed as for instance in \citep{zhang2021diffusion,gong2021interpreting}.

%
\section{Experiment}

>>>>

The structure of the transformation (e.g. Eq.~\ref{eq-NVP_forward}) is a single split of the $\bm{z}$ vector coordinates in two parts: the first part $(z_{\leq s})$ is not transformed while the transformation of the second part ${z_{>s}})$ uses a function depending only on the first part. Notice that the conditioner functions do depends only on $(z_{\leq s})$ for both the forward or the backward transformations. This single split schema is illustrated on Figure \ref{fig-nvp}. 
%
\begin{figure}
\centering
\includegraphics[width=0.45\columnwidth]{fig-nvp.pdf}
\includegraphics[width=0.45\columnwidth]{fig-nvp-inverse.pdf}
\caption{Affine coupling layer schema of forward and backward directions (Eqs.~\ref{eq-NVP_forward},\ref{eq-NVP_backward}) with a single split mechanism (figure inspired by \cite{Papamakarios2021}).}
\label{fig-nvp}
\end{figure}
%
This splitting schema implemented in a so-called \textit{coupling layer} makes the flow-based model computationally efficient and using deep neural networks, one can achieve complex $(\bm{\alpha},\bm{\beta})$ functions. Now, designers of complete architectures can compose different \textit{coupling layers} by reversing $\bm{z}$ elements between layers to ensure that all elements are transformed at the end of the set of operations. An generalisation of any permutation ordering of the inputs can be performed using a $1\times 1$ convolution layer. \cite{2016arXiv160508803D} improves the modelling using a multi-scale approach which would be prohibitive to describe here, and \cite{Kingma2018} use all these extensions to build the \texttt{Glow} model whose schema is displayed on Figure \ref{fig-Glow-archi}.
\begin{figure}
    \centering
    \vspace{-0cm}
    \begin{subfigure}[b]{0.45\textwidth}
    \begin{center}
        \centering
        \includegraphics[width=.7\textwidth]{fig-Glow-1.pdf}
        \caption{One step of Glow flow.}
    \end{center}
    \end{subfigure}%
    \vspace{5mm}
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \includegraphics[width=.99\textwidth]{fig-Glow-2.pdf}
        \caption{Multi-scale architecture \citep{2016arXiv160508803D}.}
    \end{subfigure}
    \caption{Figure from \citep{Kingma2018}. The flow step is composed of a scale and bias layer with data dependent initialization (\texttt{actnorm}), a $1\times 1$ invertible convolution layer and an affine coupling layer with $(\bm{\alpha},\bm{\beta})$ issued from a shallow convolutional neural network. Both $\bm{x}$ and $\bm{y}$ are tensors of
shape $[h \times w \times c]$ with spatial dimensions $(h, w)$ and channel dimension $c$ ans $(i, j)$ denote
spatial indices into tensors $\bm{x}$ and $\bm{y}$.  This flow step is embedded in a multi-scale  architecture which has $K$ flow steps and a number of levels $L$ \citep{2016arXiv160508803D}. The squeezing operation reduces a $s \times s \times c$ tensor  into a $s/2\times s/2 \times 4c$ tensor which mix spatial and channel components. At each scale, each channel image is divided into $2\times 2$ patches ($s=2$) and each patch is squeezed. }
    \label{fig-Glow-archi}
\end{figure}
 %
\subsection{Score based diffusion models}
%
To generate samples from $p(\bm{x})$, we can proceed as follows \citep[e.g][]{Chang2023,LinYang2023}. Starting from a sample $\bm{x}_0$ assumed to be drawn from the distribution $p_0(\bm{x})=p(\bm{x})$, we progressively transform it until the  distribution is as simple as for instance $\mathcal{N}(\bm{z},\bm{0},\bm{1})$.  The idea is then to reverse the process (the \textit{backward} mode) by generating a new sample from a new instance of noise. This transfer mapping looks very similar in concept as normalizing flows. Their latent spaces have the same dimension as the data space, there is no compression as for VAE architecture, but the main difference consists in the fact that the normalizing flows are deterministic by essence while the \textit{score diffusion} models are stochastic.

@@

Notice that in a VAE architecture,  the transformation from $\bm{x}_0$ to $\bm{z}$ is performed by an \textit{encoder} whose implementation is typically a deep neural networks with trained parameters. This is not the case in the above stochastic process where only noise is added to the original signal (e.g. galaxy image). In the same spirit, the VAE \textit{decoder} or the inverted flows are replaces by stochastic differential equation (damped Langevin)
\begin{equation}
dx_{T-t} = (x_{T-t} + 2\nabla_x \log p_{T-t}(x_{T-t}))dt + \sqrt{2}\ dB_t \quad t\in[0,T]
\label{eq-backward-diffusion}
\end{equation}
where $s_t=\nabla_x \log p_{T-t}(\bm{x}_{T-t})$ is the \textit{score} of the probability density at time $T-t$  of the blurred image $x_{T-t}$ by white noise density of  variance $\sigma^2_{T-t}$. This backward process is the generative direction to get $\bm{x}$ samples from $\mathcal{N}(\bm{z},\bm{0},\bm{1})$ sampling.

The key problem for the generation is how to obtain the \textit{score}. In this context, one may think of two possible approaches: either one gets a $p_t$ model that can exploit regularity patterns and assumptions about correlations to provide a constrained parametric framework for learning the \textit{score} by \textit{score matching} \citep{hyvarinen2005a}, or one uses a \textit{denoising} neural network as introduced by \cite{Bengio2013} and then \cite{Sohl-Dickstein2015,Ho2020} and recently studied from a mathematical point of view by \citep{kadkhodaie2024generalization}. The former approach has been used by \cite[e.g.][]{Guth2022b,Lempereur2024} using Gibbs energy parametrisation, and the later approach as been used in astrophysics for instance for some applications as the generation of galaxy images \citep{smith2021}, the generation of 21~cm luminosity temperature maps \citep{Zhao2023}, super-resolved large-scale cosmic structure predictions \citep{Schanz2023} and sampling of the high-dimensional Bayesian posterior of the weak lensing mass-mapping problem \citep{Remy2023}.

The relation between a perfect \textit{denoiser} model and the \textit{score} has been pointed by several authors \cite[e.g.][]{tweedie1947functions,herbert1956empirical,miyasawa1961empirical} and more recently by \cite{Raphan2011} for Gaussian white noise $\mathcal{N}(0,\sigma^2)$ independent of the unnoisy signal $x$. Independently of the prior on $\bm{x}$, the following theorem applies: the estimator noted $\tilde{\bm{x}}$ of the signal $\bm{x}$, such that the noisy signal is $\bm{x}_\sigma=\bm{x}+\bm{z}$ with $\bm{z}\sim\mathcal{N}(0,\sigma^2)$, is given by
\begin{equation}
\tilde{\bm{x}}= \bm{x}_\sigma + \sigma^2 \nabla_x \log p(\bm{x}_\sigma)
\label{thm-score-denoising}
\end{equation}
%
So, the use of a \textit{denoiser} network in a generative model is the schematically the following. First, we blindly train the model to recover original images from blurred versions by Gaussian white noise whose variances is not fixed \textit{a priori} and not known by the network and has the same range (e.g $[0,1]$) as the image one. Then, at each step $t$ of the backward stochastic process described above, we use the \textit{denoiser} to obtain the denoised estimate ($\tilde{\bm{x}}$) of the image $\bm{x}_{T-t}$ ($\bm{x}_\sigma$) blurred by a white noise of known variance $\sigma^2=\sigma^2_{T-t}$, so one gets the score $s_t$ to perform one step further in the backward direction.

Denoising diffusion stochastic models have gain in popularity due to the fact that they get as good as results as GAN and have certain advantages. Notably they do not need any adversarial networks, the mathematical ground is clear and robust, the training is also theoretically well defined. But they use an iterative schema for the generation that can be computationally intensive and slower compared to GAN, and the implementation of stochastic processes may require some theoretical expertise even if this argument can be mitigated. More discussion about the comparison between the two architectures  can be found in this article \cite{dhariwal2021diffusion}. We do not investigate further the mixing of flow-based and diffusion-based models as \citep[e.g.][]{zhang2021diffusion,gong2021interpreting} as well as the use of diffusion models to improve GAN stability as introduced by \cite{Wang2022}. This can be envisaged in a future study.
%
\section{Experiment}
%
Having briefly outlined different architectures of generative models in the previous section it turns out that one aspect in common between diffusion-based and flow-based models consists in the generation process of a $\bm{x}$ sample which starts by sampling a centred isotropic Gaussian distribution $\bm{z}\sim \mathcal{N}(\bm{z},\bm{0},\bm{1})$ (i.e. a white noise) of dimension $d$ equal to the final sample. But a question rises: if we train two generative models of the same kind (i.e, two flow-based, or two diffusion-based) and if we feed them with the same white noise image, does the generative process of the two models give the same final image? If not we can ask ourselves about the nature of the learned process and if we are faced to a real probability density $p(\bm{x})$ sample, or for instance does the result is a sort of mixture of the original dataset images? An answer has been proposed in \cite{kadkhodaie2024generalization} and point out size of the dataset in the passage from \textit{memorization} to \textit{generation} regime. We propose in the next section to replay their experiment using the same \texttt{U-Net} \textit{denoiser} architecture but trained with small sized ($64\times 64$) normalized gray level (1 channel $[0,1]$) centred galaxy images and in comparison use also the \texttt{Glow} flow-based architecture (Fig.~\ref{fig-Glow-archi}).
